Short-term synaptic and neuronal plasticity act in synergy for deviance detection in spiking networks

### 2022-12-12 paper-1/manuscript.tex
Rev sec Results.TA (F4)

To understand why the same stimulus A evoked a higher response in dev than in msc, we first compared the average responses in the two conditions. As shown in Figure \ref{fig4}a, we found that the response developed in a qualitatively similar fashion, with a tendency of dev spikes to occur earlier, particularly shortly after stimulation, and more frequently, particularly towards the later part of the response. Since the stimulus was identical in both conditions, this difference had to be due to either changed firing thresholds, or changed synaptic currents as a consequence of activity differences in earlier parts of the response. We quantified the former with the strength of adaptation $V_{TA}$ (Figure \ref{fig4}b), which showed a clear and widespread lowering of thresholds at the start of dev trials relative to msc, followed by an increase beyond the msc baseline in many neurons as they spiked. Correlating the difference in $V_{TA}$ at the beginning of trials with the difference in trial spike count, we found that, across neurons, lower thresholds are strongly predictive of higher activity (r = -0.424, p = 3.1e-36, see Figure \ref{fig4}d).

To quantify the synaptic contributions to a neuron's membrane potential $v$, we calculated a quantity $V_{syn}$, shown in Figure \ref{fig4}c, which captures all synaptic inputs, but excludes voltage deflections caused by direct stimulation and by the self-inhibition induced by resetting to $v_{reset} < v_{rest}$ after a spike. Being induced by spiking activity, this measure closely followed the activity pattern, making the earlier activation of the dev response clearly visible across the network. In addition, since $V_{syn}$ decays with the membrane timescale $\tau_m$, but resets to $0$ on postsynaptic spikes, many of the neurons that spiked more frequently showed lower $V_{syn}$ towards the end of network activity.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{Responses and threshold adaptation in a TA-only ablation. \textbf{a, b, c} Post-stimulus histograms across target trials, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev and msc sequence, respectively; right column, contrast between these two. \textbf{d} Per-neuron relationship between the contrast (dev - msc) in $V_{TA}$ at the start of trials, and the contrast in average number of spikes fired, excluding inhibitory neurons, and associated linear regression (r = -0.424, p = 3.1e-36). \textbf{e} Contribution of $V_{TA}$ and $V_{syn}$ to bins with increased firing in dev (i.e., red bins in the spike probability contrast, \textbf{a}), weighted by that same contrast. \textbf{f} Contribution of $V_{TA}$ relative to the sum of the weighted contributions in \textbf{e}. \textbf{g} Relative contribution of $V_{TA}$ across networks, showing median (solid line) and inter-quartile range (shaded area)}
    \label{fig4}
\end{figure*}

\textbf{Fig. 4} Responses and threshold adaptation in a TA-only ablation. \textbf{a, b, c} Post-stimulus histograms across target trials, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev and msc sequence, respectively; right column, contrast between these two. \textbf{d} Per-neuron relationship between the contrast (dev - msc) in $V_{TA}$ at the start of trials, and the contrast in average number of spikes fired, excluding inhibitory neurons, and associated linear regression (r = -0.424, p = 3.1e-36). \textbf{e} Contribution of $V_{TA}$ and $V_{syn}$ to bins with increased firing in dev (i.e., red bins in the spike probability contrast, \textbf{a}), weighted by that same contrast. \textbf{f} Contribution of $V_{TA}$ relative to the sum of the weighted contributions in \textbf{e}. \textbf{g} Relative contribution of $V_{TA}$ across networks, showing median (solid line) and inter-quartile range (shaded area)

Based on these measures, we then asked why there were more spikes in dev trials. To answer this, we weighted the contrasts in $V_{TA}$ and $V_{syn}$ by the spike probability contrast, discarding any instances of equal or lower firing in dev, and summed over neurons. The resulting time courses (Figure \ref{fig4}e) reveal the immediate underlying causes of network-wide activity increases, namely de-adaptation and depolarization. We found that increased depolarization induced by synaptic input greatly outweighed the impact of reduced adaptation throughout most of the response. Only in the very early stages of the response did adaptation dominate the increase in firing (Figure \ref{fig4}f), whereas the late response was enlarged primarily as a result of increased presynaptic excitation. This trend was reflected in most networks investigated, as shown in the summary plot in Figure \ref{fig4}g. To summarize, while most added spikes arose from increased prior activity within a trial, the initial impetus -- the increase in the early response -- was largely caused by lower TA.

>>>>

To understand why the same stimulus A evoked a higher response in dev than in msc with TA only, we first compared the average responses in the two conditions, sorting neurons by their response onset to permit a high-resolution view of the population activity. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both conditions. The contrast (right) revealed a tendency of dev spikes to occur earlier, particularly shortly after stimulation, as shown by the succession of orange (higher dev activity) by blue (higher msc activity) in the lower parts of the panel. In later parts of the response (top half of the panel), the dev response was larger, but not obviously shifted in time. Since the stimulus was identical in both conditions, and TA was the only plsaticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

We quantified the former with the strength of adaptation $V_{TA}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of dev trials relative to msc, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). In contrast, once the bulk of activity had occurred, many neurons showed higher thresholds in dev (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $V_{TA}$.

To assess the difference in synaptic currents in a way that allowed direct comparison to $V_{TA}$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs (resets after spikes notwithstanding). As the direct cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in dev around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses and threshold adaptation with TA only.
\textbf{a} Target trial average post-stimulus spike histogram, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with axes, neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with axes, neuron order and columns as in panel a.
\textbf{d} Relationship across excitatory neurons between the contrast (dev - msc) in $V_{TA}$ at the start of target trials, and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Median contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (p < 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (t = -17.3, p = 2.57e-57, median -0.166 mV; one-sided t-test, n=800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson correlation, $\rho$ = -0.424, p = 3.1e-36). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Median $V_{TA}$ was lower in dev than msc (t = -11.2, p = 1.08e-20, median = -0.094 mV; one-sided t-test, n=120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (t = -23.1, p = 3.72e-46, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two conditions. In order to assess the relative contribution of $V_{TA}$ to the increased dev response, we chose to focus exclusively on time points with higher firing probability in the dev sequence, discarding data points with equal or lower firing probability relative to the msc sequence. We then weighted the remaining contrasts in $V_{TA}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased dev response. While thresholds were slighly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $V_{TA}$ constituted a major driver of increased firing only right at the onset of the response, with later contributions amounting to less than 20\% of the total voltage difference. This pattern was consistent across networks and stimuli, as shown in Figure \ref{fig4}h. This indicates that adaptation initiated and dominated the increase in firing in the beginning, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.
